---
output:
  word_document: default
  html_document: default
  pdf_document: default
---
#Title: Final project proposal

#Description: 
#Data cleaning.
#Data analysis.

#Hypothesis: 
#Null: Immunization factors in developing countries have the same significance 
# in determining life expectancy levels compared to developed countries. 
#Alt: Immunization factors in developing countries have a greater significance in 
# determining the life expectancy level compared to developed countries. 
#Problem to solve: What factors should be targeted when trying to increase 
# the life expectancy across developing and developed countries. 

#Immunization factors: 

# 1. Hepatitis B (not significant)
# 2. Measles (not significant)
# 3. Polio
# 4. Diphtheria
# 5. HIV/AIDS

# Approach: 
# Divide the dataset in developed and developing countries, run preliminary regression
# to indeitify significant variables, check the p-values/coefficients for each
# immunization factor variable. 

# Libraries --------------------------------------------------------------------

#Probably will have to install some of these libraries. 

```{r}
install.packages('tidyr')
install.packages('psych')
install.packages('QuantPsyc')
install.packages('reshape')
install.packages('reshape2')
install.packages('MOTE')
install.packages('Hmisc')
install.packages('mice')
install.packages('corrplot')
install.packages('cocor')
install.packages('pwr')
library(tidyr)
library (psych)
library(QuantPsyc)    
library(ggplot2)		
library(moments)        
library(reshape)     	
library(reshape2)
library(MOTE)     		
library(mice)         		 
library(Hmisc)
library(ppcor)				
library(corrplot)   		
library(cocor)       	       
library(pwr)                    
options(scipen = 999)
```
# Loading file -----------------------------------------------------------------

```{r}
path = ("/Users/pratik/Desktop/Harrisburg University programs/Courses/Late Spring Courses 2022/ANLY 500/Project/Final Project Submission/R script and data/Life Expectancy Data.csv")

data = read.csv(path, header = TRUE) #change file name is applicable

View(data)
```
# Data Cleaning ================================================================

# 1. Data accuracy-----------------------------------------------------------------

```{r}
str(data)
summary(data)

data$Status = factor(data$Status)
```
#data = data %>% 
#  filter(Year %in% c(2010, 2011, 2012, 2013, 2014))    #Notfiltering last 5 years

```{r}
data = subset(data, subset = (data$Year >= 2010 & data$Year <= 2014)) # filtering from 2010 to 2014
View(data)
summary(data)
```

```{r}
install.packages('cocor')
library(cocor)
correlations = cor(data[ , -c(1,2,3)], use = "pairwise.complete.obs")
symnum(correlations)
corrplot(correlations)
```

# 2. Missing data -----------------------------------------------------------------

```{r}
miss_data_percentage = function(x){
  sum(is.na(x))/length(x)*100
}
```

```{r}
missing_rows = apply(data,1,miss_data_percentage)
missing_rows
```

```{r}
table(missing_rows)
```


```{r}
replace_rows = subset(data, missing_rows > 0 | missing_rows <= 45)
no_replace = subset(data, missing_rows > 45)
replace_rows
```

```{r}
NROW(data)
NROW(replace_rows)
NROW(no_replace)
NROW(replace_rows)+NROW(no_replace)
```

```{r}
missing_2 = apply(data[ , -c(1,2,3)], 2, miss_data_percentage)
missing_2
table(missing_2)
```

#We will exclude the categorical columns
```{r}
replace_column = replace_rows[ , -c(1,2,3)]
no_replace_column = replace_rows[ , c(1,2,3)]
```

#replacing missing data
```{r}
library(mice)
temp_no_miss = mice(replace_column, method = 'cart')

nomiss = complete(temp_no_miss, 1)

all_columns = cbind(no_replace_column, nomiss)

all_rows = rbind(no_replace,all_columns)

dim(all_columns)
```

```{r}
dim(all_rows)
```

```{r}
str(all_rows)
summary(all_rows)
```

# 3. Outliers

# We will run a fake regression model to detect outliers, excluding Categorical 
# variables from the dataset 

```{r}
output = lm(Life.expectancy ~ Adult.Mortality + infant.deaths + Alcohol + percentage.expenditure + 
Hepatitis.B + Measles + BMI + under.five.deaths + Polio + Total.expenditure + Diphtheria + HIV.AIDS + GDP + Population + 
thinness..1.19.years + thinness.5.9.years + Income.composition.of.resources + Schooling,  data = all_rows)
summary(output)
```

# 3.1.Leverage
```{r}
k=18. # k = Number of predictors, n = number of observations, which is nrow(all_rows)
leverage = hatvalues(output)
cutleverage = (2*k+2) / nrow(all_rows) #same dataset for nrow as used for output
cutleverage
```


```{r}
badleverage = as.numeric(leverage > cutleverage)
badleverage
table(badleverage)
```
# It appears there are 92 values which are above the cutleverage score, which makes them as outliers for Leverage

# 3.2 Cooks
```{r}
k=18
cooks = cooks.distance(output)
#cutcooks = 4 / (nrow(all_rows) - k - 1) 
cutcooks = 4 / (nrow(all_rows) - k -1)
cutcooks
```

```{r}
badcooks = as.numeric(cooks > cutcooks)
table(badcooks)
```
# It appears there ate 62 values which are above the cutleverage score, which makes them as outliers for Cooks

#3.3. Mahalanobis
```{r}
mahal = mahalanobis(all_columns[ ,-c(1,2,3)],
                    colMeans(all_columns[,-c(1,2,3)], na.rm = TRUE),
                    cov(all_columns[,-c(1,2,3)], use = "pairwise.complete.obs"),
                    tol=1e-20)

cutmahal = qchisq(0.999, ncol(all_columns[ , -c(1,2,3)]))
ncol(all_columns[ , -c(1,2,3)])
cutmahal
```

```{r}
badmahal = as.numeric(mahal > cutmahal)
table(badmahal)
``` 
# It appears there are 75 observations which are above the cutoff score which makes them as outliers for Mahalanobis

# 3.4 Total outliers
```{r}
totalout = badmahal + badleverage + badcooks
table(totalout)
```

# As we see that there are 50 outliers common to two conditions 29 outliers common to three, we can say that we have 
#79 (50+29) outliers.
```{r}
no_outliers_data = subset(all_rows, totalout < 2)

dim(no_outliers_data)
```

```{r}
str(no_outliers_data)
```
# 4 Assumptions ==================================================================
# We run fake regression again, but this time on the no_outliers_data dataset by removing outliers and the Population columns to test for assumptions
```{r}
output1 = lm(Life.expectancy ~ Adult.Mortality + infant.deaths + Alcohol + percentage.expenditure + Hepatitis.B + Measles + BMI + under.five.deaths + Polio + Total.expenditure + Diphtheria + HIV.AIDS + GDP + Population  + thinness..1.19.years + thinness.5.9.years + Income.composition.of.resources + Schooling, data = no_outliers_data)
str(output1)
summary(output1)
```


# Additivity -------------------------------------------------------------------
```{r}
correl = cor(no_outliers_data[,-c(1,2,3)], use = "pairwise.complete.obs")
correl
symnum(correl)
corrplot(correl)

summary(output1, correlation = T)
```

# We have 2 cases of multicollinearity and 2 cases of Singularity
#Singularity
   #S.1 Between columns Under.five.deaths and infant.deaths
      #S.1.1 We will exclude the Infant deaths column to avoid Singularity
  #S.2 Between columns thinness..1.19.years and thinness.5.9 years.
      #S.2.2 We will exclude the column thinness.5.9 years to avoid Singularity

# Multicollineriaty
    #M.1  Between columns GDP and percentage.expenditure
        # M.1.1 We will exclude the percentage.expenditure column to avoid Multicollinearity
    #M.2 Between Schooling and Income.composition.of.resources 
     # M.2.2 We will exclude the Income.composition.of.resources column to avoid 
     
     
```{r}
final_data = no_outliers_data[ , -c(6,8,20,21)]
str(final_data)
```

#Running Fake Regression after excluding the above columns
```{r}
output2 = lm(Life.expectancy ~ Adult.Mortality + Alcohol + Hepatitis.B + Measles + BMI + under.five.deaths + Polio + Total.expenditure + Diphtheria + HIV.AIDS + GDP + Population + thinness..1.19.years + Schooling, data = final_data)
summary(output2)

```

# Linearity 
```{r}
  standardized = rstudent(output2)
  fitted = scale(output2$fitted.values)
  str(fitted)
  qqnorm(standardized)
  abline(0,1)
```
#It appears that the residuals are centered over 0 with most residuals been between -2 and 2 and almost Linear  aroundthe centre. Hence we can say that we have met the assumption for Linearity.


# Normality --------------------------------------------------------------------
```{r}
hist(standardized, breaks = 20)
```

```{r}
install.packages("moments")
library(moments)
skewness(final_data[ , -c(1,2,3)], na.rm = TRUE)
```
# The Histogram of the standardized residuals however look normally distributed, with most data close to the centerand spread between -2 and 2. Some of the non-signifcant predictors like Measles and Population are skewed, but this is accpetable as they are not impacting the model been non-significantHence we can say that we have met the assumption for normality


```{r}
kurtosis(final_data[ , -c(1,2,3)], na.rm = TRUE)
```
#The distribution looks slightly kurtotic but this is acceptable as most data is centred over 0 and between -2 and 2.

# Homogeneity/Homoscedasticity -------------------------------------------------
```{r}
plot(fitted, standardized)
abline(0,0)
abline(v = 0)
```
#It appears that the residuals are equally spread above and below the X axis and all the way along the X-axis.Hence we can say that we have met the assumption for Homogeniety and Homoscadescticity.


# Final real regression

```{r}
model = lm(Life.expectancy ~ Adult.Mortality + Alcohol + Hepatitis.B + Measles + BMI + under.five.deaths + Polio + Total.expenditure + Diphtheria + HIV.AIDS + GDP + Population + thinness..1.19.years + Schooling, data = final_data)
summary(model)
```


```{r}
modelall = lm(Life.expectancy ~., data = final_data)
summary(modelall)
```

#Significance of the Overall model.We can say that the overall model is significant as the effect size(R-squared value) is very high at 0.81, the p-value is less than 0.00001 for a high f-statistic of 264.4.

#Significance of Individual predictors:
#It appears from the model that  Hepatitis B, Measles, Polio and population are the only non-significant predictors in our model since they have a p-value more than 0.05 and a small t-statistic. 
#Schooling seems to be the strongest and the strongest and the most significant positive predictor to predictLife expectancy followed by total expenditure, Alcohol, GDP and Diphtheria in that order. 
#HIV.AIDS is the strongest negative predictor for Life expectancy followed  by Adult Mortality, Under five deaths,thinness 1 to 19 and Hepatitis B.
#Alcohol and Diphtheria were shown as positive predictors for Life Expectancy which was an interesting find.
#For the significant predictors for every one point increase in the variable we get b points increase in the value of 
#y.Schooling is the strongest positive and most significant predictor for Life Expectancy

```{r}
library(ggplot2)
cleanup = theme(panel.grid.major = element_blank(),
                panel.grid.minor = element_blank(),
                panel.background = element_blank(),
                axis.line = element_line(color = "black"),
                legend.key = element_rect(fill = "white"),
                text = element_text(size = 15))
fitted1 = model$fitted.values
str(fitted1)
library(ggplot2)
scatter = ggplot(final_data, aes(fitted1, Life.expectancy))
summary(scatter)
scatter + cleanup
scatter + cleanup + geom_point() + geom_smooth(method = "lm", color = "black") + xlab("Adult.Mortality + Alcohol  + 
Hepatitis.B + Measles + BMI + under.five.deaths + Polio + Total.expenditure + Diphtheria + HIV.AIDS + GDP + 
Population + thinness..1.19.years + Schooling,  data = final_data") + ylab("Life.expectancy")
```

# t-test and effect size
```{r}
names(final_data)
cleandata = final_data[, -c(1,2)]
str(cleandata)
names(cleandata)
```

```{r}
library(dplyr)
developed = cleandata %>% filter(Status == "Developed")
str(developed)
```
```{r}
table(developed$Status)
```


```{r}
notdeveloped = cleandata %>% filter(Status ==  "Developing")
str(notdeveloped)
```

```{r}
table(notdeveloped$Status)
```

```{r}
mean(developed$Life.expectancy)

mean(notdeveloped$Life.expectancy)
```


```{r}
sd(developed$Life.expectancy)

sd(notdeveloped$Life.expectancy)
```

```{r}
length(developed$Life.expectancy)

length(notdeveloped$Life.expectancy)
```

# Two sample t-testing
```{r}

t.test(Life.expectancy ~ Status, data = cleandata,
       var.equal = T,
       paired = F)
```

```{r}
install.packages("MOTE")
library(MOTE)
effect_size = d.ind.t(m1 = mean(developed$Life.expectancy),
                      m2 = mean(notdeveloped$Life.expectancy),
                      sd1 = sd(developed$Life.expectancy),
                      sd2 = sd(notdeveloped$Life.expectancy),
                      n1 = length(developed$Life.expectancy),
                      n2 = length(notdeveloped$Life.expectancy),
                      a = 0.05)

effect_size$d
```

# Hierarchical Regression for developed countries:
# Removing "Status"(variable with index 1) as status will have only one level for each subset
# which is "Developed" for developed subset and "developing for "notdeveloped", 
# and Regression model needs atleast 2 levels for factor variables.
```{r}
developed2 = developed[,-1]
notdeveloped2 = notdeveloped[,-1]
str(developed2)
```

```{r}
str(notdeveloped2)
```

##Checking the number of uniques in "developed" dataset to ensure no factorial avriables with only one level,
## as this is important for Regression
```{r}
sapply(lapply(developed2, unique), length)
```

```{r}
model_developed = lm(Life.expectancy ~. , data = developed2) 
summary(model_developed)
```

##Checking the number of uniques in "notdeveloped" dataset to ensure no factorial avriables with only one level,
## as this is important for Regression
```{r}
sapply(lapply(notdeveloped2, unique), length)

```
## Removing "Status" as status has only one level which is "Developed", and 
## Regression model needs atleast 2 levels for factor variables.
```{r}
model_notdeveloped = lm(Life.expectancy ~. , data = notdeveloped2)
summary(model_notdeveloped)
```

# Includiong only the significant variables  as they are less and Target variable which is "Life.expectancy"
```{r}
developed2 = developed2 %>% 
  subset(select = c(Life.expectancy,
                    Adult.Mortality,
                    GDP,
                   thinness..1.19.years
  ))
```


# Removing the non significatn variables as they are less, dataset
```{r}
notdeveloped2 <- notdeveloped2 %>% 
  subset(select = -c(Hepatitis.B,
                     Measles,
                     Polio,
                     Diphtheria,
                     Population,
                     thinness..1.19.years
  ))

```

```{r}
model_developed2 = lm(Life.expectancy ~ ., data = developed2)
summary(model_developed2)
```

```{r}
model_notdeveloped2 = lm(Life.expectancy ~ ., data = notdeveloped2)
summary(model_notdeveloped2)
```
## Annova for developed (all variables) and final developed2(only significant for developed)
```{r}
anova(model_developed,
      model_developed2)
```

## Annova for notdeveloped (all variables) and final notdeveloped2(only significant for developed)
```{r}
anova(model_notdeveloped,
      model_notdeveloped2)
```

# END --------------------------------------------------------------------------
